Live Functional Programming with Typed Holes

### 2018-04-16 calculus.tex

The syntax of the core calculus, specified in Fig.~\ref{fig:hazelnut-live-syntax}, consists of types and expressions with holes. 
We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

%% Separately, we are integrating the mechanisms developed in this paper
%% into the \Hazel programming environment being developed
%% by \citet{HazelnutSNAPL} with the goal of realizing a full-scale
%% hole-driven development experience rooted in the type theoretic
%% foundations established by \Hazelnut.
%% %
%% Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper. 
%% For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
%% We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
%% \todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
%% The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}

%% Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}

@@

%The result of evaluation is a final internal expression with hole closures, each with an associated hole environment, $\sigma$. These hole environments can be reported directly to the programmer, e.g. via the sidebar shown in Fig.~X\todo{fig}, to help them as they think about how to fill in the corresponding hole in the external expression. Hole environments might also be useful indirectly, e.g. by informing an edit action synthesis and suggestion system. In any case,

When the programmer interactively fills a program hole, that program's
reduction and result are generally each affected, and the system
should respond by (quickly) updating the program's output.
%
Na\"ively, this response requires recomputing the program output
``from scratch'' after each such edit.
%
However, by virtue of its design principles (based on
CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
evaluation} of previously indeterminate expressions, which are
(generally) more determined after the programmer fills a hole than
beforehand.
%
This \emph{fill-and-resume} behavior formalizes the (ad hoc)
``edit-and-resume'' behavior permitted by some current systems,
e.g.,~Microsoft Visual Studio.\todo{cite}

>>>>

The syntax of the core calculus specified in Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with holes. 
We distinguish between \emph{external} expressions, $e$, and \emph{internal} expressions, $d$. 
External expressions correspond to programs as entered by the programmer 
(see Sec.~\ref{sec:intro}\todo{or maybe Sec. 2?}{} for discussion of manual, semi-automated and fully automated hole entry methods). 
Each well-typed external expression (as specified in Sec.~\ref{sec:external-statics} below) expands to a well-typed internal expression (see Sec.~\ref{sec:expansion}) before it is evaluated (see Sec.~\ref{sec:evaluation}). 
We take this approach, notably also taken in the specification of Standard ML by \citet{Harper00atype-theoretic}, because (1) the external language supports type inference and explicit type ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate ascriptions and specify a type assignment system when defining the dynamic semantics; and 
(2) we need additional syntactic machinery during evaluation for tracking hole closures and dynamic type casts. 
This machinery is inserted by the expansion step, rather than entered explicitly by the programmer. 
In this regard, the internal language is analogous to the cast calculus in the gradually typed lambda calculus \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the \HazelnutLive internal language goes beyond the cast calculus in several respects. We have mechanized these formal developments using the Agda proof assistant \cite{norell:thesis,norell2009dependently} 
(see Sec.~\ref{sec:agda-mechanization}). We have also implemented \HazelnutLive in a manner that maintains a novel end-to-end well-definedness invariant (see Sec.~\ref{sec:implementation}).

@@

@@

%The result of evaluation is a final internal expression with hole closures, each with an associated hole environment, $\sigma$. These hole environments can be reported directly to the programmer, e.g. via the sidebar shown in Fig.~X\todo{fig}, to help them as they think about how to fill in the corresponding hole in the external expression. Hole environments might also be useful indirectly, e.g. by informing an edit action synthesis and suggestion system. In any case, 
% When the programmer performs one or more edit actions to fill in a hole in the program, a new result must be computed. Na\"ively, the system would need to compute the result ``from scratch'' on each such edit. A more efficient approach supported by the structure of \HazelnutLive is to resume evaluation from where it left off after an edit that amounts to hole filling. We call this feature \emph{fill-and-resume}, by analogy to the ``edit-and-resume'' features available in some systems, e.g. Visual Studio (see below for a comparison)\todo{cite and compare}{}.

\todo{cite tanimoto paper, mention Level 4 liveness property}When the programmer interactively fills a program hole, that program's
reduction and result are generally each affected, and the system
should respond by (quickly) updating the program's output.
%
Na\"ively, this response requires recomputing the program output
``from scratch'' after each such edit.
%
However, by virtue of its design principles (based on
CMTT), \HazelnutLive offers a more efficient approach: To \emph{resume
evaluation} of previously indeterminate expressions, which are
(generally) more determined after the programmer fills a hole than
beforehand.
%
This \emph{fill-and-resume} behavior formalizes the (ad hoc)
``edit-and-resume'' behavior permitted by some current systems,
e.g.,~Microsoft Visual Studio.\todo{cite}
